---
title: "Basic Statistics"
output: html_document
date: "2023-06-11"
---

## Basic Statistics

R was originally developed as a statistical programming language and is commonly
used to perform basic statistics. There are also many community developed packages
that make it easy to perform statistical analyses. This tutorial will cover
some statistical tests that might be used on environmental data.

First we load the `chicago_air` ozone data so that we can run some basic 
statistics on it. We will give the ozone column its own variable name so that we
can access it quickly.


```{r, eval=FALSE}
library(region5air)
data(chicago_air)
ozone <- chicago_air$ozone

```

```{r, echo=FALSE}
load("../data/chicago_air.rda")
ozone <- chicago_air$ozone
```

### Descriptive and exploratory statistics

There are several built-in functions for descriptive statistics. The functions
below all have a parameter `na.rm` that tells the function to remove `NA` values
when performing the calculation.

```{r}

min(ozone, na.rm=T)

```

```{r}

max(ozone, na.rm=T)

```

```{r}

range(ozone, na.rm=T)

```

```{r}

summary(ozone)

```

The `hist()` and `boxplot()` functions plot a histogram and boxplot respectively,
which helps to visualize the distribution of the values.

```{r}

hist(ozone)

```

```{r}

boxplot(ozone)

```


### Measures of central tendency

R has functions for finding the mean and median of a set of values.

```{r}

mean(ozone, na.rm=T)

```

```{r}

median(ozone, na.rm=T)

```


### Measures of dispersion

The functions `var()`, `sd()`, and `IQR()` calculate the variance, standard 
deviation, and interquartile range respectively.

```{r}

var(ozone,na.rm=T)

```

```{r}

sd(ozone,na.rm=T)

```

```{r}

IQR(ozone,na.rm=T)

```

## Useful packages for statistics


```{r}

# install.packages("Hmisc")
library(Hmisc)
describe(ozone)

```
```{r, warning=FALSE, message=FALSE}
# install.packages("psych")
library(psych)
describeBy(ozone)

```


```{r}
# install.packages("pastecs")
library(pastecs)

stat.desc(ozone, norm=T) 

```


## Statistical tests

R has many built-in functions for statistical tests. As an example, we'll use
the `t.test()` function to perform a two sample t-test on the Chicago ozone data.

First, let’s plot our dataset and look at ozone values by month.

```{r, warning=FALSE, message=FALSE}

library(ggplot2)
ggplot(chicago_air, aes(factor(month), ozone)) + geom_boxplot()

```
We could compare ozone months in March and June and see if there is a significant
difference in concentrations. 


```{r, warning=FALSE, message=FALSE}

ozone_march_june <- chicago_air[chicago_air$month %in% c(3, 6), ]

ggplot(ozone_march_june, aes(factor(month), ozone)) + geom_boxplot()

```


We should also check for normality before doing any statistical tests.


```{r}

ggplot(ozone_march_june, aes(ozone)) + 
  facet_grid(rows = "month") + 
  geom_histogram()

```


d.mar <- density(March.data$ozone)
d.aug <- density(August.data$ozone,na.rm=T)
plot(d.mar, main="Kernel Density Plot of ozone data")
par(new=TRUE)
plot(d.aug, xlab="", main="", xaxt="n", yaxt="n")


Now we can do some comparisons between these 2 months of readings using the student’s t test.

t.test(March.data$ozone,August.data$ozone)
## 
##  Welch Two Sample t-test
## 
## data:  March.data$ozone and August.data$ozone
## t = -3.7499, df = 43.053, p-value = 0.0005233
## alternative hypothesis: true difference in means is not equal to 0
## 95 percent confidence interval:
##  -0.012861130 -0.003865751
## sample estimates:
##  mean of x  mean of y 
## 0.03890323 0.04726667
t.test results show that these means are in fact different.
We may be curious as to how each of the months compares with each other. We can do a pairwise.t.test to look at all the months.
# this takes the form pairwise.t.test(x,g), where x = the vector of values and g= the grouping vector or factor.
pairwise.t.test(chicago_air$ozone,chicago_air$month)  
## 
##  Pairwise comparisons using t tests with pooled SD 
## 
## data:  chicago_air$ozone and chicago_air$month 
## 
##    1       2       3       4       5       6       7       8       9      
## 2  0.15321 -       -       -       -       -       -       -       -      
## 3  1.7e-08 0.01310 -       -       -       -       -       -       -      
## 4  1.8e-15 2.6e-07 0.26779 -       -       -       -       -       -      
## 5  < 2e-16 2.4e-11 0.00174 1.00000 -       -       -       -       -      
## 6  < 2e-16 3.1e-12 0.00041 1.00000 1.00000 -       -       -       -      
## 7  1.00000 1.00000 0.39405 0.00536 0.00014 5.7e-05 -       -       -      
## 8  < 2e-16 4.0e-10 0.00857 1.00000 1.00000 1.00000 0.00041 -       -      
## 9  5.7e-11 0.00039 1.00000 1.00000 0.05957 0.01853 0.09282 0.19596 -      
## 10 1.00000 0.90415 1.1e-06 3.4e-13 < 2e-16 < 2e-16 1.00000 < 2e-16 5.7e-09
## 11 1.00000 0.00645 4.1e-11 < 2e-16 < 2e-16 < 2e-16 1.00000 < 2e-16 9.1e-14
## 12 1.00000 0.00526 2.2e-11 < 2e-16 < 2e-16 < 2e-16 1.00000 < 2e-16 4.3e-14
##    10      11     
## 2  -       -      
## 3  -       -      
## 4  -       -      
## 5  -       -      
## 6  -       -      
## 7  -       -      
## 8  -       -      
## 9  -       -      
## 10 -       -      
## 11 1.00000 -      
## 12 1.00000 1.00000
## 
## P value adjustment method: holm
Correlation analysis
If we want to look at how all the variables in our dataset relate to each other, correlation analysis is very useful.

To run a correlation test, all of the values you give it must be numeric; therefore, dates or character values won’t work.
We need to subset our data to just the numeric columns that we want to test for correlation.
You will want to save the results to a variable so you can use it for plotting later.
chi_air.cor <- cor(chicago_air[,2:5], 
                   use= "complete.obs", 
                   method="pearson") #subset to only numeric data, can choose between spearman, kendall, and pearson correlation coefficients
chi_air.cor  #print data to console to see the correlation matrix
##           ozone      temp     solar     month
## ozone 1.0000000 0.6035925 0.5926064 0.3397222
## temp  0.6035925 1.0000000 0.4923545 0.8138164
## solar 0.5926064 0.4923545 1.0000000 0.4385685
## month 0.3397222 0.8138164 0.4385685 1.0000000
cor.test(chicago_air$solar, chicago_air$temp, method = "pearson")
## 
##  Pearson's product-moment correlation
## 
## data:  chicago_air$solar and chicago_air$temp
## t = 9.4859, df = 254, p-value < 2.2e-16
## alternative hypothesis: true correlation is not equal to 0
## 95 percent confidence interval:
##  0.4148735 0.5966489
## sample estimates:
##       cor 
## 0.5114604
A package called corrplot has some really nice plotting options for correlation matrices.

install.packages("corrplot")
library(corrplot)
corrplot(chi_air.cor, order='FPC')  #you can order the plots based on alphabetical order, first principal component order, hierarchical clustering order, and more

corrplot(chi_air.cor, 
         method='number', 
         order='FPC')


Pairwise plots
pairs(chicago_air[,2:5], lower.panel=panel.smooth)  # simply does pairwise plotting for all numeric variables

library(psych)

pairs.panels(chicago_air[,2:3], lm = TRUE)  # shows the bivariate scatterplots, histogram, and Pearson correlation for each pair.
pairs.panels(chicago_air[,2:5], lm = TRUE)


Reference Table of a few popular tests for categorical data analysis in R
test	function
Chi Square Test	chisq.test()
Fisher’s Test	fisher.test()
Analysis of Variance	aov()
The EnvStats package has a comprehensive list of basic and more advanced statistical tests for Environmental Data.
library(EnvStats)
?EnvStats
Now let’s try some exercises to test our understanding of basic statistical analysis in R.