---
title: "Basic Statistics"
output: 
  html_document: 
    keep_md: yes
date: "`r Sys.Date()`"
---

R was originally developed as a statistical programming language and its built-in
functions are commonly used for basic statistics. There are also many community 
developed packages that make it easy to perform statistical analyses. This tutorial
will cover descriptive statistics functions and some statistical tests that might
be used on environmental data.


- [Prerequisites](#prerequisites)
- [Descriptive Statistics](#descriptive-statistics)
- [Statistical Tests](#statistical-tests)
- [The envstats package](#the-envstats-package)
- [Exercises](#exercises)

<br>
<hr>
<br>

# Prerequisites

This lesson assumes you are familiar with the material in the lesson on
[Functions and Importing Data](../2-Functions-and-Importing-Data/readme.md).

Statistical functions are used in this lesson that require installation of the
following packages. 

```{r, eval=FALSE}
install.packages("envstats")
```

The data from the R package `region5air` is used throughout these lessons.
To install the package from GitHub, use the `remotes` package. Run the code
below to install the `remotes` package and install `region5air` from GitHub.

```{r, eval=FALSE}
# if you have not installed remotes
install.packages("remotes")

library(remotes)
install_github("FluentData/region5air")
```

To load the `chicago_air` data frame we will be using in the lesson, use the 
`library()` function to load the `region5air` package, then the `data( )` function
to load the data frame. 

```{r}

library(region5air)

data(chicago_air)

```



# Descriptive Statistics

R has many built-in functions for descriptive statistics. We will use these 
functions to understand the ozone data in the `chicago_air` data frame.

```{r}

ozone <- chicago_air$ozone

```

Most of the functions we'll be using have an argument named `na.rm` that stands 
for `NA` remove. If the argument is set to `TRUE` then the function will remove
all missing values from the data set. Otherwise the function will error.

These functions let us know the range of the ozone values, i.e. the highest and 
lowest values.

```{r}

min(ozone, na.rm=TRUE)

```

```{r}

max(ozone, na.rm=TRUE)

```

```{r}

range(ozone, na.rm=TRUE)

```

We can also get the range and the quartile values from the `summary()` function.

```{r}

summary(ozone)

```

The `1st Qu.` value is the first quartile. This means that a quarter of the ozone
values are between the minimum value `r min(ozone, na.rm = TRUE)` and the first
quartile value `r quantile(ozone, .25, TRUE)`. The `Median` value is the second
quartile, meaning half the values are between the minimum and the second quartile
value of `r median(ozone, na.rm = TRUE)`.

```{r}

IQR(ozone, na.rm=TRUE)

```

The `hist()` and `boxplot()` functions plot a histogram and boxplot respectively,
which helps to visualize the distribution of the values.

```{r}

hist(ozone)

```

```{r}

boxplot(ozone)

```


### Measures of central tendency

R has functions for finding the mean and median of a set of values.

```{r}

mean(ozone, na.rm=T)

```

```{r}

median(ozone, na.rm=T)

```


### Measures of dispersion

The functions `var()`, `sd()`, and `IQR()` calculate the variance, standard 
deviation, and interquartile range respectively.

```{r}

var(ozone,na.rm=T)

```

```{r}

sd(ozone,na.rm=T)

```

```{r}

IQR(ozone,na.rm=T)

```

## Useful packages for statistics

Below are some helpful functions that provide statistical descriptions from 
the packages `Hmisc`, `psych`, and `pastecs`.

```{r, warning=FALSE, message=FALSE}

# install.packages("Hmisc")
library(Hmisc)
describe(ozone)

```

```{r, warning=FALSE, message=FALSE}
# install.packages("psych")
library(psych)
describeBy(ozone)

```


```{r, message=FALSE, warning=FALSE}
# install.packages("pastecs")
library(pastecs)

stat.desc(ozone, norm=T) 

```


## Statistical tests

R has many built-in functions for statistical tests. As an example, we'll use
the `t.test()` function to perform a two sample t-test on the Chicago ozone data.

First, let’s plot our dataset and look at ozone values by month.

```{r, warning=FALSE, message=FALSE}

library(ggplot2)
ggplot(chicago_air, aes(factor(month), ozone)) + geom_boxplot()

```

We could compare ozone months in March and June and see if there is a significant
difference in concentrations. 


```{r, warning=FALSE, message=FALSE}

ozone_march_june <- chicago_air[chicago_air$month %in% c(3, 6), ]

ggplot(ozone_march_june, aes(factor(month), ozone)) + geom_boxplot()

```


We should also check for normality before doing any statistical tests. Below
are histograms of the datasets.


```{r}

ggplot(ozone_march_june, aes(ozone)) + 
  facet_grid(rows = "month") + 
  geom_histogram()

```


We can also perform a Shapiro-Wilk test using the `shapiro.test()` function.

```{r}

ozone_march <- ozone_march_june[ozone_march_june$month == 3, "ozone"]

shapiro.test(ozone_march)
```

```{r}

ozone_june <- ozone_march_june[ozone_march_june$month == 6, "ozone"]

shapiro.test(ozone_june)
```

The p-value for the tests are well above 0.05, so it fails to reject the null
hypothesis. Meaning, we can assume the distribution of ozone in the two months
are normal.

Now we can do some comparisons between these 2 months of readings using the 
student's t test.

```{r}

t.test(ozone_march, ozone_june)

```


The `t.test()` output shows that these means are in fact different.

## Correlation analysis

If we want to look at how all the variables in our dataset relate to each other,
correlation analysis is very useful.

To run a correlation test, all of the values you give it must be numeric; therefore, 
dates or character values won’t work. We need to subset our data to just the 
numeric columns that we want to test for correlation. You will want to save the 
results to a variable so you can use it for plotting later.

First we make a `data.frame` that is a subset of the numeric columns of our
dataset.

```{r}

chicago_numeric <- chicago_air[, c("ozone", "temp", "solar")]

```

Then we use the `cor()` function to print a correlation matrix, specifying in the
arguments that we only want to include complete observations (no records with 
missing values) and the Pearson method of finding correlations.


```{r}

cor_matrix <- cor(chicago_numeric, 
                  use = "complete.obs", 
                  method ="pearson") 
 
cor_matrix

```

To perform a correlation test, we use the `cor.test()` function.

```{r}

cor.test(chicago_air$solar, chicago_air$temp, method = "pearson")

```

The `ggcorrplot` package uses `ggplot2` to create a correlation matrix graph.

```{r}

library(ggcorrplot)

ggcorrplot(cor_matrix, lab = TRUE)

```


## Pairwise plots

It's also useful to see pairwise plots for numeric values. There is a built-in
`pairs()` function in R, as well as a `pairs.panels()` function from the
`psych` package.

```{r}
pairs(chicago_numeric, lower.panel=panel.smooth)  

```

```{r}
library(psych)

pairs.panels(chicago_numeric, lm = TRUE) 

```

## Other statistical tests

Below is a reference table of a few popular tests for categorical data analysis in R.

test |	function |
---|---|
Chi Square Test	| chisq.test()
Fisher’s Test	| fisher.test()
Analysis of Variance |	aov()


The `EnvStats` package has a comprehensive list of basic and more advanced statistical tests for Environmental Data.

```{r, eval=FALSE}
# install.packages("EnvStats")

library(EnvStats)

?EnvStats

```