---
title: "Subsetting and Sorting Data"
output: 
  html_document: 
    keep_md: yes
date: "`r Sys.Date()`"
---

# Working with R `data.frame`s

We'll begin by loading the `chicago_air` `data.frame` from the `region5air`
[package on GitHub](https://github.com/NateByers/region5air).

```{r, eval=FALSE}
library(region5air)

data(chicago_air)
```

```{r, echo=FALSE}
load("../data/chicago_air.rda")
```

We always want to make sure our data looks the way it is supposed to before we 
begin working with it.

The best way to take a quick look at the first few rows of a data frame
is to use the `head()` function

```{r}
head(chicago_air)  
```
You can specify the number of lines to display by using the n = parameter

```{r}
head(chicago_air, n = 3)
```

You can also look at the bottom of the data frame by using `tail()`

```{r}
tail(chicago_air)
```

The table function is helpful for summarizing your data by counts and the `plot()`
and `hist()` functions allow you to quickly visualize the data

```{r}
table(chicago_air$ozone)  ##Summarizes by counts
plot(chicago_air$ozone)  # Quick plot of data
hist(chicago_air$ozone)  #Like a historgram plot except no binning occurs
```


# Indexing

If we want to look at a small subset of the data, you can subset by using the 
index values of the `data.frame` to grab particular rows and columns. An index 
value is just like reading coordinates on a map, where the format is `[rows, columns]`.
Below is an example of how you access a particular value in a `data.frame` based on its index.

```{r}
my_data <- data.frame(x = 1:5, y = 6:10)
my_data[4, 2] ## This should grab the value in the 4th row and 2nd column 
```


We can also access data from a vector using the same indexing idea. In this case,
you don’t need the comma to separate the rows and columns since you are accessing 
one dimensional data.

```{r}
x <- c(1, 3, 2, 7, 25.3, 6)
x[5]  # This will access the fifth element in the vector
```


Now that we understand indexing we can subset the `chicago_air` data frame by using the 
`[` function.

# Subsetting Using Indexing
To get one row of the data frame, specify the row number you would like in the 
brackets, on the left side of the comma. By leaving the column value on the right
side of the comma blank, it returns all the columns associated with row number 1.

```{r}
chicago_air[1, ]
```


If you want more than one row, you can supply a vector of row numbers

```{r}
chicago_air[c(1, 2, 5), ] #Accesses the 1, 2 and 5th rows of data
```

To get a column from the data frame, specify the column number in the brackets, 
to the right of the comma. By leaving the row value blank, you are telling it to
return all rows associated with column 1.

```{r}
head( chicago_air[, 1] )
```

You can obtain more than one column by supplying a vector of column numbers

```{r}
head( chicago_air[, c(3, 4, 6)] )
```

Column names can also be used.

```{r}
head( chicago_air[, "solar"] )
```

Or a vector of column names

```{r}
head( chicago_air[, c("ozone", "temp", "month")] )
```

Both rows and columns can be specified at the same time. 

```{r}
chicago_air[1:5, 3:5]  # Returns first 5 rows of data and the third through fifth columns.
```

# Logical Operators
You can also subset a data frame by using logical expressions
The logical expression is used to specify rows that you want to keep or discard


## Reference Table of Logical Operators
|Operator |Description |
| :---    | :---       |
| <	      | less than  |
| <=	    | less than or equal to|
| >	      | greater than |
| >=	    | greater than or equal to |
| ==	    | exactly equal to |
| !=	    | not equal to |
| !x      | not x |
| x & y   | x AND y |
| x <code>&#124;</code> y |	x OR y|

Here is an example using a logical expression to subset.

```{r}
logical_vector <- chicago_air$solar > 1.45 # records with greater than 1.45 in the solar column

chicago_air[logical_vector, ] 
```


Another helpful tool when subsetting is the complete.cases function.
This function allows us to only look at data where observations for all columns are complete.
The function returns a logical vector that can be used to subset a `data.frame`.

```{r}
complete <- complete.cases(chicago_air)

air <- chicago_air[complete,] 
```


_We will use the `air` `data.frame` for the rest of the subsetting
examples._


Let's say we only want rows in this `data.frame` where ozone was above 70 ppb (.070 ppm).

```{r}
ozone_violation <- air[(air$ozone > .070), ]  # This returns all the days with readings above .070 ppm

ozone_violation
```


If we wanted all of the days in the 7th month, we could use the `==` operator.

```{r}
air[(air$month == 7), ]
```

Or if we want all days except the 6th day, we would use `!=`.

```{r}
head( air[(air$weekday != 6), ] ) #Excludes all data from the 6th day of the week
```

We can combine logical conditions with `&` (the _AND_ operator)

If we wanted only rows where the temperature was between 80 and 85 (including those numbers)

```{r}
air[(air$temp >= 80 & air$temp <= 85), ]
```

We can also use `|` (the _OR_ operator) to select rows on days 3 or 5
```{r}
head( air[(air$weekday == 3 | air$weekday == 5),] )
```

# Subsetting using the subset() function

You can also use the `subset()` function to filter a `data.frame` down to the records
you want. The first argument in the function is the name of the `data.frame` and
the second argument is the logical expression. No need to use `$` for column names.

```{r}
high_temp <- subset(air, temp > 90)  

head(high_temp)
```

By using the select = parameter you can specify which columns to keep
```{r}
high_temp_ozone <- subset(air, temp > 90, select = c(ozone, temp))

head(high_temp_ozone)

```



# Sorting data 

You can sort the rows of a `data.frame` using the `order()` function. For 
example, we can sort the `chicago_air` dataset by ozone. The output of the `order()`
function is a vector of integers that map to the ascending order of the input. 


```{r}
ozone_ordered <- order(air$ozone)

head(ozone_ordered)

```

We can use the output to arrange the rows of the `data.frame` by placing the 
ordered vector on the left side of the `[, ]` operator.

```{r}
air_ozone_ascending <- air[ozone_ordered, ] 

head(air_ozone_ascending)
```


You can easily sort the `data.frame` in descending order by reversing the
ordered vector, using the `rev()` function.

```{r}
air_ozone_descending <- air[rev(ozone_ordered), ] 

head(air_ozone_descending)
```

# Combining `data.frame`s

`data.frame`s can be combined using the `rbind()`  and `cbind()` functions, standing
for "row bind" and "column bind" respectively. 

The `rbind()` function requires that there to be an an equal number of columns
among the `data.frame`s. To illustrate, we will make two subsets of the `air`
`data.frame` and then combine them using `rbind()`.

```{r}

nrow(air) # show the original number of records

air_warm <- subset(air, temp > 80) # get warm air records

nrow(air_warm) # show the number of records

air_cool <- subset(air, temp <= 80) # get cool air records

nrow(air_cool) # show number of records

air_recombined <- rbind(air_warm, air_cool) # combine the data.frames

nrow(air_recombined) # number of records


```



