---
title: "Intro to R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to write functions, use control structures, and loop through data
  using for loops and apply functions.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

tutorial_options(
  exercise.timelimit = 60,
  # A simple checker function that just returns the message in the check chunk
  exercise.checker = function(check_code, ...) {
    list(
      message = eval(parse(text = check_code)),
      correct = logical(0),
      type = "info",
      location = "append"
    )
  }
)
knitr::opts_chunk$set(error = TRUE)
gradethis::gradethis_setup()
```

## Welcome



In this tutorial, you will learn how to:

* make your own custom function
* use if statements with logical expressions 
* loop through vectors and lists
* use the `apply()` function on a `data.frame`



### Prerequisites

We will use the `chicago_air` `data.frame` from the `region5air` package. Below
is code for installing the package and loading the `data.frame` into your R session.

```{r, eval=FALSE}
library(devtools)
install_github("natebyers/region5air")

library(region5air)
data(chicago_air)
```

```{r chicago_air, echo=FALSE}
load("data/chicago_air.rda")
```



## Writing functions

We already know that when we use a function, we need to know what the arguments 
are. For example, if we want to use the `mean()` function, we look at the 
documentation by typing `?mean()`.


The first function we write will be our own version of a function that calculates
the mean of a numeric vector. Since we can’t just call it mean (or, if we do, 
we will replace that base function in our environment), we will call our function
`myMean()`.

Here is the basic structure of creating a function.

```{r, eval=FALSE}
myMean <- function( ){
  
}
```

A function is created by using another function called `function()`. The 
parentheses will contain the arguments we want our function to have, and the 
curly braces will contain the operation that will be performed on the arguments.
First, we'll need a vector of numbers, so we'll call it the `x` argument.

```{r, eval=FALSE}
myMean <- function(x){
  #do something to x
}
```

Now we’ll calculate the average of the numbers in x.

```{r function1, exercise = TRUE, exercise.eval = FALSE}
my_mean <- function(x){
  
  count <- length(x)
  sum <- sum(x)
  
  average <- sum/count
  
  average
}
```

Inside the curly braces we've found how many numbers are in the x vector, we've
added up all the values in that vector, and we've found the average. In a function,
the last line will always be returned. 

Let’s try our function.

```{r function2, exercise = TRUE, exercise.eval = FALSE, exercise.setup = 'function1'}

my_vector <- c(1, 3, 5, 2, 6, 9, 0)

my_mean(x = my_vector)

```

Now let’s check our answer against the mean function that is built into R

```{r function3, exercise = TRUE, exercise.eval = FALSE, exercise.setup = 'function2'}

my_vector <- c(1, 3, 5, 2, 6, 9, 0)

mean(x = my_vector)

```

### Exercise

Write a function that converts ozone concentrations from ppm to ug/m3.
First have the function convert ppm to ppb, and then multiply ppb by 1.96 to 
account for molecular mass of O3 and molar volume. Check the function by converting
`0.078` ppm and printing the answer.

```{r exercise1, exercise = TRUE, exercise.eval = FALSE}

convert03 <- function(x) {
  
  
  
}

```

```{r exercise1-solution}
convert03 <- function(x){
  ppb <- x * 1000
  ug <- ppb * 1.96
  ug
}

convert03(0.078)
```

```{r exercise1-check}
grade_result(
  pass_if(~identical(.result, 152.88))
)
```

## Conditionals

Conditional statements are helpful if you want to write some code that will do
one thing in some circumstances and something else the rest of the time. R has 
the conditional functions `if()` and `ifelse()` that perform these operations. 
These are very similar to the `IF` function used in Excel and they take the same 
type of arguments

`if()` uses braces after the parentheses to group operations based on whether or
not the logical expression is true. Below is pseudo-code that will not run in
R, but shows an outline of how to use the `if()` function:

```{r, eval=FALSE}
if(<logical expression>) {
  
  # code that will run if the logical expression is true
  
} else {
  
  # code that will run if the logical expression is false
  
}


```

For example, if we wanted to print a message based on the value of ozone, we 
could use this construction:

```{r conditional1, exercise = TRUE, exercise.eval = FALSE}

ozone_value <- 0.075

if(ozone_value > 0.065) {
  
  print("Potential Health Effects")
  
} else {
  
  print("All Good")
  
}
```

The `ifelse()` function is a simplified version of `if()` and can be used when 
you are returning simple values based on a conditional test.

```{r, eval=FALSE}

ifelse(<test>, <yes>, <no>)

```

The `test` argument is the logical expression, `yes` is the value returned if
the expression is true, and `no` is returned if the expression is false.

```{r conditional2, exercise = TRUE, exercise.eval = FALSE}

ozone_value <- 0.060

message <- ifelse(ozone_value > 0.065, "Potential Health Effects", "All Good")

print(message)

```

### Exercise

Use the `ifelse()` function and the vector of ozone values below to create
a logical vector indicating `TRUE` if the value is greater than 0.065. Be sure
to have the results printed and submit your answer.

```{r exercise2, exercise = TRUE, exercise.eval = FALSE}

ozone <- c(0.050, 0.060, 0.070, 0.065, 0.068)


```

```{r exercise2-solution}

ozone <- c(0.050, 0.060, 0.070, 0.065, 0.068)

ifelse(ozone > 0.065, TRUE, FALSE)

```

```{r exercise2-check}
grade_result(
  pass_if(~identical(.result, c(FALSE, FALSE, TRUE, FALSE, TRUE)))
)
```

## For loop

Like most programming languages, R has for and while loops. This tutorial will
cover just for loops and move on to `apply()` functions, which are more commonly 
used in R.

For loops are used to repeat an operation a set number of times. The basic outline is

```{r, eval=FALSE}

for(i in sequence){
  
}
```

The `sequence` parameter is typically a vector. The `i` parameter is a variable
that will take on the values in the `sequence` in order. For instance, if `sequence`
was the vector `c(1, 2, 3)` then the `i` variable will take on each of those values
in turn.

This example simply prints the values of the vector as they loop through.

```{r for1, exercise = TRUE, exercise.eval = FALSE}
ints <- c(1, 2, 3)

for(i in ints) {
  
  print(i)
  
}

```

Below is a more complicated example. Three vectors are saved in a `list` variable,
and the mean of each vector is saved in a variable named `my_averages`. The 
`my_averages` variable is created first as an empty vector `c()` then populated
in the for loop. We use a vector of integers as a sequence to access the list
values and save the output in the vector.

```{r for2, exercise = TRUE, exercise.eval = FALSE}
my_averages <- c()

my_list <- list(c(1, 5, 9, 3),
                1:10, 
                c(23, 42))

for(i in c(1, 2, 3)){
  
  my_averages[i] <- mean(my_list[[i]]) 
  
}

my_averages
```

### Exercise

Use a for loop to find the maximum value for each vector in the list below, and 
save to the vector `max_values`. Be sure to print the results and submit your answer.


```{r exercise3, exercise = TRUE, exercise.eval = FALSE}

max_values <- c()

samples <- list(c(0.032, 0.037, 0.037, 0.044, 0.035, 0.041, 0.037, 0.035),
                c(0.064, 0.061, 0.058, 0.040, 0.035, 0.060, 0.060, 0.053),
                c(0.036, 0.033, 0.041, 0.035, 0.038, 0.035, 0.033, 0.033))






```

```{r exercise3-solution}

max_values <- c()

samples <- list(c(0.032, 0.037, 0.037, 0.044, 0.035, 0.041, 0.037, 0.035),
                c(0.064, 0.061, 0.058, 0.040, 0.035, 0.060, 0.060, 0.053),
                c(0.036, 0.033, 0.041, 0.035, 0.038, 0.035, 0.033, 0.033))

for(i in c(1, 2, 3)){
  
  max_values[i] <- max(samples[[i]]) 
  
}

max_values

```

```{r exercise3-check}
grade_result(
  pass_if(~identical(.result, c(0.044, 0.064, 0.041)))
)
```

## `apply()` functions

In R, the most efficient way to do loops is to use `apply()` and related functions,
such as `lapply()`, `tapply()`, and `mapply()`. 

`apply()` takes a `data.frame` (or matrix) as the first argument. The second 
argument specifies if you want to apply a function to the rows (1) or columns (2),
and the third argument is the function you want to apply to each row or column.
Additional arguments can be used to pass on to the function being applied to each row or column.

The example below applies the max function to three of the columns of the `chicago_air` 
`data.frame`. 

```{r apply1, exercise = TRUE, exercise.eval = FALSE, exercise.setup = 'chicago_air'}

air <- chicago_air[, c("ozone", "temp", "solar")]

air_max <- apply(air, 
                 MARGIN = 2,  # we are applying the max() function to each column
                 FUN = max, 
                 na.rm = TRUE # the na.rm argument is passed to the max() function
                 )
air_max
```


`lapply()` applies a function to each member of a list. Below we apply the 
mean function to each vector in a list, similar to the for loop above. The output
is also a list.

```{r apply2, exercise = TRUE, exercise.eval = FALSE}

my_list <- list(c(1, 5, 9, 3), 1:10, c(23, 42))

my_averages <- lapply(my_list, mean)

my_averages
```

`sapply()` does the same thing as `lapply()`, but the output is simplified as 
much as possible. In the example below, a vector is returned.

```{r apply3, exercise = TRUE, exercise.eval = FALSE}
my_list <- list(c(1, 5, 9, 3), 1:10, c(23, 42))

my_averages <- sapply(my_list, mean)

my_averages
```

### Exercise

Use the `data.frame` below and the `apply()` function to get the maximum value 
for each monitor. Be sure to print your result and submit your answer.

```{r exercise4, exercise = TRUE, exercise.eval = FALSE}

samples <- data.frame(
  monitor1 = c(0.032, 0.037, 0.037, 0.044, 0.035, 0.041, 0.037, 0.035),
  monitor2 = c(0.064, 0.061, 0.058, 0.040, 0.035, 0.060, 0.060, 0.053),
  monitor3 = c(0.036, 0.033, 0.041, 0.035, 0.038, 0.035, 0.033, 0.033)
)






```

```{r exercise4-solution}

samples <- data.frame(
  monitor1 = c(0.032, 0.037, 0.037, 0.044, 0.035, 0.041, 0.037, 0.035),
  monitor2 = c(0.064, 0.061, 0.058, 0.040, 0.035, 0.060, 0.060, 0.053),
  monitor3 = c(0.036, 0.033, 0.041, 0.035, 0.038, 0.035, 0.033, 0.033)
)

x <- apply(samples, MARGIN = 2, FUN = max)

str(x)
```

```{r exercise4-check}

grade_this({

    x <- c(0.044, 0.064, 0.041)
    names(x) <- c("monitor1", "monitor2", "monitor3")
    if (identical(.result, x)) {
      pass("Great work!")
    }

    
  })

```